A Simple Minibatch Acceptance Test for MCMC

### 2016-05-23 nips_2016_jfc.tex
added changes from JFC

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to
perform Metropolis-Hastings tests on the full data. To do this, we
describe a novel acceptance test which is applied to minibatches of
data. Previous work on minibatch Metropolis-Hastings use adaptive
batch sizes, which may use far more than one minibatch of data and therefore
yield few samples.  Here we describe a simpler test on fixed-sized
minibatches. Instead of adaptive batch sizes, our test requires a
precondition on the variance of the log proposal probability
ratio. Given a measurement of this variance, a minibatch size can be
chosen to support the test in one step, which yields a number of
samples proportional to the batch/minibatch ratio. Alternatively, the
variance condition can be used to adjust the temperature of the
sampling distribution to a ``natural'' value to provide regular
samples at a given minibatch size. The resulting proposal/test is only
slightly more complex than a simple SGD update. In this paper we
derive the test, discuss its implementation, and present several
experiments.
\end{abstract}

@@

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^N p(x_i \mid \theta)$ of
parameter $\theta$ based on conditionally independent data
$\{x_i\}_{i=1}^N$ is the target distribution which one aims to
approximate by sampling $\theta$ values. When $N$ is large, it is
very expensive to compute the exact posterior because the
likelihood must be evaluated on every $x_i$.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and therefore have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, MCMC methods typically
require calculations over the full dataset to produce a new sample. A
recent result from~\cite{cutting_mh_2014} attempted to bridge the gap
between minibatch optimization and MCMC by proposing an \emph{adaptive
  minibatch} method. Their method uses a subset of the $N$ data points
for the MH test each iteration. However the number of points is not
known a priori, and is adjusted dynamically during every test. This
undermines many of the practical advantages of minibatch calculations: 
GPU kernels work best for certain minibatch sizes, memory blocks
can be reused when minibatch size is fixed, etc.

@@

There is another advantage when using our test for Bayesian posterior inference. As the dataset size
grows, the posterior distribution sharpens and often concentrates to a few sharp peaks. Samples from
this distribution will come from those peaks, and will easily become stuck in one of them. The
samples are not exploring the posterior parameter space, and are unlikely to concentrate at a strong
local optimum. They therefore fail to accurately represent a sample of the posterior because of
inadequate search (i.e., mixing). We argue that it is much more natural to sample the distribution
at a higher temperature determined by the minibatch size rather than the dataset size. Starting at a
high temperature (small minibatch size) which mixes easily, one can reduce the temperature
(increasing minibatch size) until one reaches the desired target distribution temperature.  Although
this approach also involves changing minibatch size, the changes increase minibatch size
monotonically, and only a few changes are performed in the course of a sampling session.  Annealing
can also be used for MAP and ML estimation because changing temperature preserves posterior modes.

@@

\begin{enumerate}[noitemsep]
    \item We develop a new minibatch acceptance test which satisfies detailed balance.
    \item We compare performance of our new test and the adaptive sampling method.
    \item We experiment using the approach for posterior estimation.
\end{enumerate}

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the distribution
$p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated) samples
$\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the sample counts.  Each
iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This is done by drawing a uniform
random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. This $P_a$ satisfies detailed balance, so if
one samples long enough, one will arrive at a stationary distribution matching the posterior, though
a burn-in period and/or only using samples at regular intervals is often done in practice.

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on a probability ratio. If the test cannot make a decision
over a certain confidence threshold, then the minibatch size is increased and the test repeats. This
process continues until a decision. The downside of this algorithm is the need to keep incrementing
the minibatch size; in the worst case, all $N$ data points may be needed to decide on $\theta'$. A
similar approach from~\cite{icml2014c1_bardenet14} has slightly more robust theoretical properties
based on concentration inequalities, but comes at the cost of more computational time.

The work of~\cite{conf/uai/MaclaurinA14} presents another minibatch MCMC method which gets the
\emph{exact} posterior, so long as there exists a (cheap) lower bound on the likelihood for each
data point. Their approach is only a starting point, however, and it can be difficult to derive
these lower bounds in general cases. The extra auxiliary variables they introduce also hurt mixing.

There is a separate line of MCMC work on simulating the physics of a probability distribution. By
viewing random variables as particles in a system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate the best case scenario for proposals: high
quality \emph{and} distant.  HMC methods require a full gradient computation of the posterior, which
means they also face the problem of computing the likelihood at each point. Stochastic versions of
HMC~\cite{sghmc_2014,stochastic_thermostats_2014} use a minibatch of data each iteration but require
controlling the minibatch noise to avoid divergence. Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} is a similar approach which does not use the
momentum terms in HMC.  Minibatch SGHMC and Langevin Dynamics are orthogonal to our work and are
other ways of attempting to solve the same problem of MCMC methods on big data. We can combine our
methods with these results, as we demonstrate in Section~\ref{ssec:nets}.

@@

For our new MH test, we use two key values, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') (\prod_{i=1}^n p(x_i \mid
\theta'))^{\frac{N}{n}} q(\theta_t \mid \theta')}{p(\theta_t) (\prod_{i=1}^n p(x_i \mid
\theta_t))^{\frac{N}{n}} q(\theta' \mid \theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}. Thus, $\Delta$ and
$\Delta'$ are probability ratios in log space. Note that the only difference between $\Delta$ and
$\Delta'$ is that they use $N$ and $n$ data terms $x_i$, respectively, and that the latter uses an
$N/n$ scaling term. We assume $n \ll N$ (so $\Delta'$ is substantially faster to compute) and that
the $n$ data points $x_i$ represent a random minibatch without replacement.

@@

\begin{lemma}\label{lem:detailed_balance}
An acceptance function $g$ such that $g(\Delta) = \exp(\Delta) g(-\Delta )$ satisfies detailed
balance. That is, $f(\theta_t)p(\theta' \mid \theta_t) = f(\theta')p(\theta_t \mid \theta')$, where
$p(\theta_y \mid \theta_x)$ is the probability of jumping from $\theta_x$ to $\theta_y$ in our
chain.
\end{lemma}

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\Delta))^{-1}$. Straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily sample from
it using the following procedure. At any iteration with a current parameter $\theta_t$ and a
candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u \sim
{\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. We can define $X = g^{-1}(u)$ so that its CDF
is the logistic function, which follows because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of $u$ is one. Thus, the criteria to accept the candidate $\theta'$ is equivalent to
whether $\Delta > X$. Moreover, the density of $X$ (a standard logistic random variable) is
symmetric about zero, so the acceptance criteria can also be expressed as $\Delta + X>0$.

@@

With large datasets, $\Delta$ is intractable to compute. An intuitive fix is to replace $\Delta$
with the minibatch version, $\Delta'$, for our MH test. The following lemma characterizes the
distribution of $\Delta'$.

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly without replacement and has sufficiently many elements,
then the distribution of $\Delta'$ is approximately Gaussian.
\end{lemma}

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\frac{N}{n}\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Gaussian for
sufficiently large $N$ by the Central Limit Theorem.
\end{proof}

% Daniel: I decided to put the figure here, so it's closer to the text that references it.
\begin{figure}[t]
    \centering
    \includegraphics[width=1\textwidth]{mh_convolution_diagram_v2}
    \caption{
    Three examples of $X_{\rm norm}$ and $X_{\rm corr}$ distributions that convolve to form the
    standard logistic distribution. We use three standard deviation values of $X_{\rm norm}$. The
    two red (respectively, blue, green) curves form the logistic. The $y$-axis is capped at 0.5.
    This figure must be viewed in color.
    }
    \label{fig:deconvolution}
    %\vspace{-10pt}
\end{figure}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The noise means we cannot just replace $\Delta+X$ with
$\Delta'+X$ for our test, so we need to slightly change our acceptance criteria. Our insight is to
decompose $X$ as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X$. The criteria
to accept is now
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists because $X_{\rm norm}$ is an estimate of $\varepsilon$. If $X_{\rm norm}$
is exact, then we get the same MH test with $\Delta'$ as we do with the full data version $\Delta$.

It is important to understand why we use Equations~\ref{eq:deconvolution} and~\ref{eq:criteria}. The
deconvolution process allows us to use $\Delta' + X_{\rm corr}$ as a proxy for $\Delta + X$. It
works because if we know the distribution\footnote{Computing the distributions of $\varepsilon$ and
$X_{\rm norm}$ reduces to finding their standard deviations.} of $\varepsilon$ (from
Equation~\ref{eq:relationship}), then we can pretend that our $\Delta'$ is in fact $\Delta$. To
``insert'' $\varepsilon$ into our test, we deconvolve $X$ so that one of its components, $X_{\rm
norm}$, plays the part of $\varepsilon$. In light of our revised MH test, the previous discussion
raises the question: in an iteration of MCMC with current $\theta_t$ and proposed $\theta'$, how do
we know the distribution of $X_{\rm corr}$? We need the correct distribution to sample the
\emph{values} of $X_{\rm corr}$ appropriately. This procedure consists of two major steps.

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}) we make $K$
    small, around 5-10, for speed purposes, and $n$ is usually 50-200.  Using these $K$ values, we
    compute the empirical standard deviation ${\rm std}(\Delta')$.

    \item Once we have ${\rm std}(\Delta')$, we can determine the distribution of $X_{\rm corr}$
    from standard deconvolution techniques. These rely on the well-known fact that the Fourier
    transform of a convolution is the product of the individual Fourier transforms. In practice, we
    pre-compute distributions of $X_{\rm corr}$ for different ${\rm std}(\Delta')$ values to
    facilitate the sampling process.
\end{enumerate}

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, number of $\Delta'$ estimates $K$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distributions, $\{x_i\}_{i=1}^N$ values from a target distribution, and initial sample $\theta_0$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ for either posterior inference or
MAP/ML estimation.}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$ and create an empty list for $\Delta'$ estimates, $d = []$\;
    \For{$k = \{1, \ldots, K\}$}{
        Sample a random minibatch of size $m$ from the complete data\;
        Compute $\Delta'$ from this minibatch (see Equation~\ref{eq:deltas}), and add $\Delta'$ to $d$\;
    }
    Compute ${\rm std}(d)$, the standard deviation estimate of $\Delta'$ from list $d$\;
    \eIf{${\rm std}(d) \ge 1.2$}{
        Skip loop (set $\theta_{t+1}=\theta_t$) and apply variance preconditioning fix (if desired)\;
    }{
        Choose the closest $X_{\rm corr}$ distribution, and sample a value $X_c$ from it\;
        Using another random minibatch of $m$ data points, compute a \emph{new} $\Delta'$ (call it $\Delta_{\rm real}'$)\;
        \eIf{$X_c + \Delta_{\rm real}' > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old one, $\theta_{t+1} = \theta_t$\;
        }
    }
}
\caption{A description of our MH test within the MCMC algorithm.}
\label{alg:our_algorithm}
%\vspace{-10pt}
\end{algorithm}

>>>>

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets where it is impractical to perform
Metropolis-Hastings tests on the full data. We describe a novel
acceptance test which can be applied to fixed-size minibatches of
data. Previous works on reducing sample count for the
Metropolis-Hastings test require a data-dependent number of samples
which is not fixed a-priori.  They do not work on bounded-size samples
and therefore do not provide the typical practical benefits of
minibatch methods.  Our novel test requires an upper bound on the
variance of the log probability, but given this condition it can be
applied to fixed-size minibatches of data. We show via experiments
that if this condition is not met, that the earlier methods consume an
large number of samples.  Our test is extremely simple and fast,
bringing the benefits of minibatch methods to MCMC inference.
Alternatively, the variance condition can be used to adjust the
temperature of the sampling distribution to a ``natural'' value to
provide regular samples at a given minibatch size. The resulting
proposal/test is only slightly more complex than a simple SGD
update. In this paper we derive the test, discuss its implementation,
and present several experiments.
\end{abstract}

@@

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the sampling target distribution over the
parameters $\theta$ is the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^np(x_i \mid \theta)$ of
based on conditionally independent data $\{x_i\}_{i=1}^n$. When $n$ is
large, it is very expensive to compute the exact posterior because the
likelihood must be evaluated on every $x_i$.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, MCMC methods typically
require calculations over the full dataset to produce a new sample.
Recent results
\cite{cutting_mh_2014,conf/uai/MaclaurinA14,icml2014c1_bardenet14}
attempted to bridge the gap between minibatch optimization and MCMC by
data-adaptive subsampling. These methods uses a
subset of the $N$ data points for the MH test each iteration. However
the number of points is not known a priori, and is adjusted
dynamically during every test. This undermines many of the practical
advantages of minibatch calculations: GPU kernels work best for
certain minibatch sizes, memory blocks can be reused when minibatch
size is fixed, etc.

@@

There is another advantage when using our test for Bayesian
posterior inference. As the dataset size grows, the posterior distribution
sharpens and often concentrates to a few sharp peaks. Samples from
this distribution will come from those peaks, and will easily become
stuck in one of them (see Section~\ref{ssec:gaussians}).  The samples are not exploring
the posterior parameter space, and are unlikely to concentrate at a
strong local optimum. They therefore fail to accurately represent a
sample of the posterior because of inadequate search (i.e.,
mixing). We argue that it is much more natural to sample the
distribution at a higher temperature determined by the
minibatch size rather than the dataset size. This temperature is
determined in a natural way by the variance prerequisite for our
test. Starting at a high temperature (small minibatch size) which
mixes easily, one can reduce the temperature (increasing minibatch
size) until one reaches the desired target distribution temperature.
Although this approach also involves changing minibatch size, the
changes increase minibatch size monotonically, and only a few changes
are performed in the course of a sampling session.
Annealing can also be used for MAP and ML estimation because changing temperature preserves
posterior modes.

@@

\begin{enumerate}[noitemsep]
\item We develop a new acceptance test which works on fixed-size data samples, given
  an auxiliary condition. 
    \item We compare performance of our new test and adaptive sampling methods.
    \item We experiment using the approach for posterior estimation.
\end{enumerate}

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the distribution
$p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated) samples
$\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the sample counts.  Each
iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This is done by drawing a uniform
random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. The $P_a$ from Equation~\ref{eq:traditional}
satisfies detailed balance, so if one samples long enough, one will arrive at a stationary
distribution matching the posterior, though a burn-in period and/or taking every $n$th sample may be
helpful.

Unfortunately, computing $f$ requires the use of all $N$ training data
points. Moreover, it is difficult to design tests using substantially
fewer than $N$ points that also satisfy detailed balance. To reconcile
these competing objectives,~\cite{cutting_mh_2014} proposes an
adaptive minibatch MCMC algorithm which uses a sequential hypothesis
test. During each iteration, the algorithm starts with a small
minibatch of data and adds additional samples until the test either
succeeds or fails. in the worst case, all $N$ data points may be
needed in a single iteration.  A similar approach is explored in
~\cite{icml2014c1_bardenet14} but as the authors there note, the
method is ``of limited usefulness'' when the log probability variance
is high, and the method may consume all or most of the samples. The
work of~\cite{conf/uai/MaclaurinA14} presents another approach to
reducing sample size for MH testing. In their case, auxiliary
variables are introduced which allow exact sampling from the
posterior. Their method assumes a lower bound on the posterior
probability. Unfortunately, in the most challenging situations (high
variance in the log probability), this lower bound, if it exists at
all, will be much smaller than the mean probability. As a consequence,
this method will consume most of the data samples.

MCMC methods can use physics-inspired state-space
extensions. By viewing random variables as particles in a physical
system and adding virtual momentum variables, one realizes the
Hamiltonian Monte Carlo (HMC)~\cite{mcmc_hamiltonian_2010} approach.
These methods generate high-efficiency proposals with large steps and
high acceptance rates. Traditional HMC methods require a full gradient
computation for proposals, but recent extensions
HMC~\cite{sghmc_2014,stochastic_thermostats_2014} use minibatches of
data. Langevin Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} is a
similar approach which does not use the momentum terms in HMC, but
typically produces smaller steps and random-walk behavior. In both
cases, these methods avoided the expense of MH acceptance tests which
are much larger than the minibatch proposals. But because of this they
are forced to limit step size so that their approximate proposals
generate the true distribution in the limit. With our fast test,
these methods can be made more efficient by using larger
step sizes without compromising accuracy.
We demonstrate this in Section~\ref{ssec:nets}.

@@

For our new MH test, we define $\Delta$ and $\Delta'$ which are logs of transition probability
ratios:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') (\prod_{i=1}^n p(x_i \mid
\theta'))^{\frac{N}{n}} q(\theta_t \mid \theta')}{p(\theta_t) (\prod_{i=1}^n p(x_i \mid
\theta_t))^{\frac{N}{n}} q(\theta' \mid \theta_t)} \right)
\end{equation} 
where $p$ and $q$ are target and proposal distributions respectively
as in Equation~\ref{eq:traditional}. Note that the difference
between $\Delta$ and $\Delta'$ is that they use $N$ and $n$ data terms
$x_i$, respectively, and that the latter uses an $N/n$ scaling
term. The scaling term ensures that $\Delta'$ is an unbiased estimator of $\Delta$.
 We assume $n \ll N$ (so $\Delta'$ is substantially faster to
compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

\begin{lemma}\label{lem:detailed_balance}
Let $\Delta$ be defined as in Equation~\ref{eq:deltas}. Any acceptance function $g$ such that
$g(\Delta) = \exp(\Delta) g(-\Delta )$ satisfies detailed balance. That is, $f(\theta_t)p(\theta'
\mid \theta_t) = f(\theta')p(\theta_t \mid \theta')$, where $p(\theta_y \mid \theta_x)$ is the
probability of jumping from $\theta_x$ to $\theta_y$ in our chain.
\end{lemma}

@@

For our MH test, the key is that we use a different $g$, the logistic
function: $g(\Delta) = (1+\exp(-\Delta))^{-1}$. Straightforward
arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is ideal for
this task because we can easily sample from it using the following
procedure: At any iteration with a current parameter $\theta_t$ and a
candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a
uniform random variable $u \sim {\rm Unif}[0,1]$. We accept $\theta'$
if $g(\Delta) > u$, and reject otherwise. This process is equivalent
to sampling a random variable $X$ with cumulative distribution
function $F_X(x) = g(x)$, and accepting if $\Delta > X$ and rejecting
otherwise. We can define $X = g^{-1}(u)$ so that its CDF is the
logistic function, which follows because for arbitrary $X = x$, we
have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of $u$ is one. Thus, the criteria to accept the candidate $\theta'$ is equivalent to
whether $\Delta > X$. Moreover, $X$ is symmetric about zero, so the acceptance criteria can also be
expressed as $\Delta + X>0$.

@@

With large datasets, $\Delta$ is intractable to compute. An intuitive fix is to instead use the
minibatch approximation $\Delta'$. The following lemma characterizes the distribution of $\Delta'$.

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly and sufficiently large, then the distribution of $\Delta'$ is approximately Normal.
\end{lemma}

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\frac{N}{n}\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the variation in $\Delta'$
comes only from the samples in the minibatch. Since each term in the sum is IID, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Normal distribtuion as $n
\to \infty$ by the Central Limit Theorem.
\end{proof}

% Daniel: I decided to put the figure here, so it's closer to the text that references it.
\begin{figure}[t]
    \centering
    \includegraphics[width=1\textwidth]{mh_convolution_diagram_v2}
    \caption{
    Three examples of $X_{\rm norm}$ and $X_{\rm corr}$ distributions that convolve to form the
    standard logistic distribution. We use three standard deviation values of $X_{\rm norm}$. The
    two red curves convolve to form the logistic, etc. The $y$-axis is capped at 0.5 for
    readability. This figure must be viewed in color.
    }
    \label{fig:deconvolution}
    %\vspace{-10pt}
\end{figure}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' \approx \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Normal noise term $\varepsilon$.  The extra noise means we can no longer directly perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. Our insight is to
decompose $X$ as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Normally-distributed and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add to form $X$. The criteria
to accept is now
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists because $X_{\rm norm}$ is an estimate of $\varepsilon$. If $X_{\rm norm}$
is exact, then we get the same MH test with $\Delta'$ as we do with the full data version $\Delta$.

Equations~\ref{eq:deconvolution} and~\ref{eq:criteria} yield an extremely simple test: Namely,
compute the approximate log transition probability ratio $\Delta'$, add the correction random
variable, and compare with zero. It remains to be determined how, and under what conditions
we can do the decomposition~\ref{eq:deconvolution}. First we recall that the distribution
of a sum of indendent random variables is the convolution of their distributions. From equation
\cite{eq:deconvolution}, we know the distributions of both $X_{\rm norm}$ and $X$. It follows
that $X_{\rm corr}$ is the {\em deconvolution} of $X$ by $X_{\rm norm}$. Such a decomposition
need not exist. For instance, \ref{eq:deconvolution} implies that
\begin{equation}
  {\rm var}(X) = {\rm var}(X_{\rm norm}) + {\rm var}(X_{\rm corr})
\end{equation}
so we must have ${\rm var}(X) > {\rm var}(X_{\rm norm})$ to have a chance of success. In fact
since these distributions are known, the variance information is enough to determine whether
the test is applicable. The correction sample can be generated as follows:

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}), $K$ is
    small, around 5-10, and $n$ is usually a few hundred for the Central Limit Theorem to apply.
    Using these $K$ values, we use the empirical standard deviation ${\rm std}(\Delta')$.

    \item Once we have ${\rm std}(\Delta')$, we can determine the distribution of $X_{\rm corr}$
    from standard deconvolution techniques. These rely on the well-known fact that the Fourier
    transform of a convolution is the product of the Fourier transforms. In practice, we pre-compute
    distributions of $X_{\rm corr}$ for different ${\rm std}(\Delta')$ values to facilitate the
    sampling process.
\end{enumerate}

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, variance estimation window $W$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distributions, $\{x_i\}_{i=1}^n$ values from a target distribution, and initial sample $\theta_0$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$ and create an empty list for $\Delta'$ estimates, $d = []$\;
    \For{$k = \{1, \ldots, W\}$}{
        Sample a random minibatch of size $m$ from the complete data\;
        Compute $\Delta'$ from this minibatch (see Equation~\ref{eq:deltas}), and add $\Delta'$ to $d$\;
    }
    Compute ${\rm std}(d)$, the standard deviation estimate of $\Delta'$ from list $d$\;
    \eIf{${\rm std}(d) \ge 1.2$}{
        Skip loop (set $\theta_{t+1}=\theta_t$) and apply variance preconditioning fix (if desired)\;
    }{
        Choose the closest $X_{\rm corr}$ distribution, and sample a value $X_c$ from it\;
        Using another random minibatch of $m$ data points, compute a \emph{new} $\Delta'$ (call it $\Delta_{\rm real}'$)\;
        \eIf{$X_c + \Delta_{\rm real}' > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the sample, $\theta_{t+1} = \theta_t$\;
        }
    }
}
\caption{A description of our MH test within the MCMC algorithm.}
\label{alg:our_algorithm}
%\vspace{-10pt}
\end{algorithm}

\textbf{The Convolution}. Figure~\ref{fig:deconvolution} demonstrates
Equation~\ref{eq:deconvolution} and provides three (color-coded) examples of densities for $X_{\rm
norm}$ and $X_{\rm corr}$. The density of $X$, a logistic random variable with mean $\mu = 0$ and
scale $s=1$, is known, as are the $X_{\rm norm}$ densities for different standard deviations. The
deconvolution provides us with the $X_{\rm corr}$ distributions. Note, however, that as the standard
deviation of $X_{\rm norm}$ increases, the $X_{\rm corr}$ distribution becomes increasingly unstable
and ``bumpier'', because the logistic curve has fatter tails than the normal distribution. For this reason, we cap
the standard deviation possibilities of $X_{\rm norm}$ at $\approx 1.2$ in our experiments. In our
experiments, we discretize the possible standard deviations of $X_{\rm norm}$, and we also limit the
densities to consider the range $[-10,10]$ (as shown in Figure~\ref{eq:deconvolution}).
